---
title: Problem set 6 
date: 2024-10-25
execute: 
  eval: false
---

Please answer each of the exercises below. For those asking for a mathematical calculation please use LaTeX to show your work.

Important: Make sure that your document renders in less than 5 minutes.

1. Write a function called `same_birthday` that takes a number `n` as an argument, randomly generates  `n` birthdays and returns `TRUE` if two or more birthdays are the same. You can assume nobody is born on February 29. 

Hint: use the functions `sample`, `duplicated`, and `any`.

```{r} 
same_birthday <- function(n){ 
  ## Your code here
} 
``` 

2. Suppose you are in a classroom with 50 people. If we assume this is a randomly selected group of 50 people, what is the chance that at least two people have the same birthday? Use a Monte Carlo simulation with $B=$1,000 trials based on the function `same_birthday` from the previous exercises. 

```{r} 
B <- 10^3
## Your code here
``` 

3. Redo the previous exercises for several values on `n` to determine at what group size do the chances become greater than 50%. Set the seed at 1997.


```{r} 
set.seed(1997)
compute_prob <- function(n, B = 10^3){ 
 ## Your code here
} 
## Your code here
``` 

4. These probabilities can be computed exactly instead of relying on Monte Carlo approximations. We use the multiplication rule:



$$
\mbox{Pr}(n\mbox{ different birthdays}) = 1 \times \frac{364}{365}\times\frac{363}{365} \dots \frac{365-n + 1}{365} 
$$


Plot the probabilities you obtained using Monte Carlos as a points and the exact probabilities with a red line. 

Hint: use the function `prod` to compute the exact probabilities.

```{r} 
exact_prob <- function(n){ 
 ## Your code here
} 
## Your code here
``` 

5. Note that the points don't quite match the red line. This is because our Monte Carlos simulation was based on only 1,000 iterations. Repeat exercise 2 but for `n = 23` and try `B <- seq(10, 250, 5)^2` number iterations. Plot the estimated probability against `sqrt(b)`. Describe when it starts to stabilize in that the estimates are within 0.005 for the exact probability. Add horizontal lines around the exact probability $\pm$ 0.005. Note this could take several seconds to run. Set the seed to 1998.

```{r}
set.seed(1998)
B <- seq(10, 250, 5)^2
## Your code here
```

6. Repeat exercise 4 but use the the results of exercise 5 to select the number of iterations so that the points practically fall on the red curve.

Hint: If the number of iterations you chose is too large, you will achieve the correct plot but your document might not render in less than five minutes.

```{r}
n <- seq(1,60) 
## Your code here
```


7. In American Roulette, with 18 red slots, 18 black slots, and 2 green slots (0 and 00), what is the probability of landing on a green slot?

$$
\mbox{Derivation here}
$$


6. The payout for winning on green is \$17 dollars. This means that if you bet a dollar and it lands on green, you get \$17. If it lands on red or black you lose your dollar. Create a sampling model using `sample` to simulate the random variable $X$ for the Casino's winnings. 

```{r}
n <- 1
## Your code here
```

7. Now create a random variable $S$ of the Casino's total winnings if $n = $1,000 people bet on green. Use Monte Carlo simulation to estimate the probability that the Casino loses money. 

```{r}
n <- 1000
## Your code here
```

8. What is the expected value of $X$? 

$$
\mbox{Your derivation here.}
$$


9. What is the standard error of $X$?

$$
\mbox{Your derivation here.}
$$


10. What is the expected value of $S$? Does the Monte Carlo simulation confirm this?

$$
\mbox{Your dereviation here}
$$

```{r}
## Your code here
```


11. What is the standard error of $S$? Does the Monte Carlos simulation confirm this?

$$
\mbox{Your derivation here.}
$$
```{r}
## Your code here
```

12. Use data visualization to convince yourself that the distribution of $S$ is approximately normal. Make a histogram and a QQ-plot of standardized values of $S$.
The QQ-plot should be on the identity line.

```{r}
## Your code here
```

13. Notice that the normal approximation is slightly off for the tails of the distribution. What would make this better? Increasing the number of people playing $n$ or the number of Monte Carlo iterations $B$?

**Answer here**

14. Now approximate the probability estimated using CLT. Does it agree with the Monte Carlo simulation?

$$
\mbox{Your derivation here.}
$$

```{r}
## Your code here
```

15. How many people $n$ must bet on green for the Casino to reduce the probability of losing money to 1%. Check your answer with a Monte Carlo simulation.

$$
\mbox{Your derivation here.}
$$
```{r}
## Your code here
```

