---
title: Problem set 10
date: 2024-12-16
execute: 
  eval: false
---

The data for this problem set is provided by this link: `https://github.com/datasciencelabs/2024/raw/refs/heads/main/data/pset-10-mnist.rds`

Read this object into R. For example, you can use:

```{r}
fn <- tempfile()
download.file("https://github.com/datasciencelabs/2024/raw/refs/heads/main/data/pset-10-mnist.rds", fn)
dat <- readRDS(fn)
file.remove(fn)
```


The object is a list with two components `dat$train` and `dat$test`. Use the data in `dat$train` to  develop a machine learning algorithms to predict the labels for the images in the `dat$test$images` component.

Save the your predicted labels in an object called `digit_predictions`. This should be a vector of integers with length `nrow(dat$test$images)`. It is important that the `digit_predictions` is ordered to match the rows of `dat$test$images`.

Save the object to a file called `digit_predictions.rds` using:

```{r}
saveRDS(digit_predictions, file = "digit_predictions.rds")
```

You will submit:

1. The file `digit_predictions.rds`

2. A quarto file that reproduces your analysis and provides brief explanations for your choices. 

**If your code reproduces the result**, your grade will be your accuracy  rounded up the closest integer. So,for example, if your accuracy is .993 your grade will be 100%. 

You will have two opportunities to submit your predictions and see your accuracy before your submitting your final predictions.



